Data Management \\ and LSST Special Programs

### 2023-05-01 ap_docrev.tex
update

Below are described mentions of Special Programs in Rubin documentations, with a focus on requirements and specifications.

Updates to Rubin documents related to Special Programs -- motivated by past versions of this DMTN -- were made via LCR-1309 and LCR-2265.

@@

The above requirement indicates that the LSST system shall be able to process non-standard visits from Special Programs, but that the image quality might be degraded.
Improvements that require algorithms or processing outside of what the LSST Science Pipelines can provide would be left to the science community and require User-Generated pipelines and data products.

@@

In Section 1.3.1.3, LSR-REQ-0124\lsrreq{0124}, the discussion specifies that an image quality parameter related to ellipticity applies only to main survey data.

@@

Version 19.1 (revision 2021-07-30), \citeds{LSE-30}.

@@

Section 3.1.5.1.2, ``Data Products Handling for Special Programs", OSS-REQ-0392\ossreq{0392} is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121), and specifies that {\it ``the handling of data products from Special Programs shall be compliant with the approach defined in LSE-163"}.

In Section 3.2.5.3, OSS-REQ-0403\ossreq{0403} is a flow-down of LSR-REQ-0124\lsrreq{0124} related to the ellipticity correlation function distribution.

In Section 3.6.1.3, ``Continuous Exposures", OSS-REQ-0319\ossreq{0319} requires that {\it ``The Observatory shall be capable of continuous operation throughout a night with the interval between successive visits equal to the FPA readout time"}.
The discussion clarifies that {\it ``this mode of observing is needed to support observations when the telescope is not being re-pointed"}, such as deep drilling fields or other Special Programs.

@@

In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field"}.
There do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits.

@@

In Section 3.6.3.8, ``Guiding", OSS-REQ-0305\ossreq{0305} specifies that {\it ``the LSST system shall improve the tracking of the optical system to the sky by means of active guiding over the whole accessible sky"}.

@@

%%% MLG: this does not appear in OSS V13
% $\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs). {\bf JIRA ticket DM-12579 confirmed that there is no maximum number, and so many Special Programs will be able to be included in the scheduler}.

%%% MLG: this does not appear in OSS V13
% $\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (OSS 2.1.1.7, Visit Optimization).

%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
% $\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively. These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin. These estimates appear adequate for Special Programs in general.

%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
% $\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively. These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.

@@

%%% MLG: the following are all from <2018 and were not updated in 2022

% \subsubsection{Extreme-Depth CoAdds} The system has been sized to hold $\sim200$ exposures in memory at once, which defined by the current maximum number of visits per field in the WFD main survey in $10$ years (from a conversation with K.-T.~Lim). Note that the panchromatic CoAdds would be built from the individual filter CoAdds, so the algorithm does not need to handle $\sim800$ images. From a computational standpoint, $200$ is the maximum number of images that can be stacked with an algorithm that requires all images to be accessible in memory at once (i.e., loading all images and calculating the median for each pixel). Deeper stacks might be possible with algorithms that deal with images consequentially. It is conceivable that a Special Program which needs to stack $>200$ images is not possible to accomplish with reconfigured pipelines, and would have to be processed with external, user-contributed resources. However, the exact DM capabilities in this area are not yet well known because NCSA has not yet defined the machine capabilities. Furthermore, the planned commissioning data will go to a $\sim20$ year depth, and so it can reasonably be expected that DM will have to be able to accommodate at least a stack that deep.

% \subsubsection{Deblending} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey. It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal. Level 3 deblenders for specific Special Programs fields may require development by the user community.

% \subsubsection{Variability Characterization} The periodic and aperiodic variability characterizations described in Section 6.21 of the DMAD are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types. From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals. It is conceivable that the goals of a particular Special Program might require different algorithms; these could be provided by DM, or written as Level 3 and either made joinable to the DM reconfigured data products or perhaps incorporated directly.

% \subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP \texttt{Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM. It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.

@@

%%% MLG commented out the paragraphs below in 2022, they were out of date.

% $\bullet$ The database schema for \texttt{DIASource} does not appear to have an element that identifies which template image that was used. This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data. DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61). In conversation with K.-T., we find that this isn't a problem, as it will be handled by provenance: the code configuration used and the time of processing are sufficient to identify and regenerate the template image. However, K.-T. has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable. However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.

% $\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs. It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this. Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType. Indicates which input was used to produce a given object."} Is that all we need? {\bf JIRA ticket DM-12580 clarified that  \texttt{Object.prv\_inputID} in the database schema is one possible way to identify whether an \texttt{Object} is an externally provided coordinate}.

% $\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
% \texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
% \texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
% Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey. Are 32 floats in each of the 6 filters always going to be a large enough volume? The Transients and Variable Stars Science Collaboration currently has a Task Force assigned to address this, and {\bf it is the topic of JIRA ticket DM-12581}.

%%% MLG commented out this whole section in 2022, since LSE-180 is from 2013
%%%  LSE-180 is deprecated?

% \subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}

% LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey. Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}. Regarding the reduction and calibration of non-standard visit images, LSE-180 makes two relevant points: \\
% $\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
% $\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD. The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.

% However, Lupton's recent work on calibrations has made much of LSE-180 obsolete, and it is not clear whether it is still needed as a separate document. At the time of writing, Lupton's most recent take on calibrations can be found in \url{https://github.com/lsst-dm/calibration/blob/master/calibration.pdf}. K.-T. referenced the Appendix B for the overview of calibration types, but it seems that there is not yet a final plan that can be assessed for its suitability for Special Programs. It is plausible that Special Programs could have their own calibration database, and obtaining additional calibration frames is not expected to cause a bottleneck (at the moment there is $\sim1.7$ hours per day for this, and exceeding this could eat into the engineering time). The most likely cause for trouble involving calibrations and Special Programs is the computational time needed to create the additional and/or different calibration files to be applied to non-standard Special Programs data, and/or the additional overhead required to load them in the Level 1 pipeline if the schedule interleaves SP-WFD-WFD-SP.

% {\bf JIRA ticket DM-12582 is currently open, and aims to define the potential additional calibration needs of Special Programs data.}

>>>>

This review is limited to requirements related to Special Programs processing or data products,
or requirements that constrain the diversity of images (e.g., exposure time limits).

Updates to Rubin documents related to Special Programs that were motivated by past versions of this DMTN
were made via LCR-1309 and LCR-2265.

@@

@@

@@

Version 19.3 (revision 2022-12-12), \citeds{LSE-30}.

@@

Section 3.1.5.1.2, ``Data Products Handling for Special Programs", OSS-REQ-0392\ossreq{0392} is a flow-down of 
requirements from the LSR (0122, 0075, and 0121; LSE-29), and specifies that {\it ``the handling of data products 
from Special Programs shall be compliant with the approach defined in LSE-163"}.

In Section 3.6.1.3, ``Continuous Exposures", OSS-REQ-0319\ossreq{0319} requires that {\it ``The Observatory shall be 
capable of continuous operation throughout a night with the interval between successive visits equal to the FPA 
readout time"}.
The discussion clarifies that {\it ``this mode of observing is needed to support observations when the telescope is 
not being re-pointed. For example observing ``deep drilling" fields..."}.

@@

In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able 
to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on 
a single field"}.

@@

@@

@@

@@
